Using numerical modelling to investigate temperature and wave exposure as drivers

### 2020-11-21 thesis_chapter_2.Rmd
intro updated to latest

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], form the basis of one of the most productive ecosystems globally [@Mann1973].  Kelps generally have a dependence on cool-temperate and arctic seawater temperatures [@Bolton2010; @Mohring2014-bk; @Cavanaugh2011-ib; @Dayton1985], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and sub-canopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Steneck2002], each depending on a wide suite of ecological services provided by kelp forests [@Gaines1987-bg; @Paul1993-bm; @Levin1994-qf; @Willis2003-xo; @anderson1997; @Dayton1985]. Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in their distribution [@Gorman2013-nj], abundance [@Cavanaugh2011-ib; @Dayton1998-xw], diversity [@Wing2007-tv; @Wernberg2008], composition [@Leliaert2000a; @Dayton1985; @Norderhaug2012-ks; @Harley2012], growth [@Cousens1982-ck] and productivity [@Pedersen2012-vr; @Dayton1998-xw; @Krumhansl2012].

Given that kelps are cool-temperate organisms which are vulnerable to dislodgement under high wave exposure scenarios, increases in storm frequency and magnitude, and changes in ocean temperature, changes in these influences pose a direct threat to their survival; which is a concern considering that these variables are known to be affected by climate change[@Russo2014; @Reguero2019; @Weisse2010; @Rose2012; @Rummukainen2012]. Warmer temperatures reduces the resilience of kelp individuals to storm disturbance [@Wernberg2010b], causes fragmentation through weakening tissues [@Simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist across a broad range of environments which is largely due to their morphological plasticity [@FowlerWalker2006; @Wernberg2010a]. A study by @Wernberg2003-mp investigated the morphology of *Ecklonia radiata* (C. Agardh) J. Agardh. in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australian coast. They found no correlation between spatial distance and morphological similarity; rather the morphology of the kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001] and effects of wave dampening [@Kobayashi1993-os; @Mia2013; @Dubi1995].

Although multiple environmental factors play a role in kelp morphological adaptation, wave exposure and temperature have been identified as the main drivers across various kelp species [@bekkby2014; @Wernberg2005c; @Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004; @mabin2013]. For example, a study by @Serisawa2002-dc compared the morphology of *Ecklonia cava* Kjellman growing at warmer and cooler sites, and showed wrinkling of the blade and reduced size to be a characteristic of higher temperature ranges. Warmer temperature affects important kelp physiological processes such as photosynthesis and respiration, which influences growth and productivity and ultimately leads to a reduction in size [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *E. radiata* across a broad geographic range, and showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of exposure. Cartographical models of wave exposure are 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitative, reproducible approach for characterising the hydrodynamic environment. Although @mabin2013 investigated kelp morphological characteristics using a 2D WAM model, currently no studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modelling. Furthermore no studies have identified specific wave metrics that drive kelp morphological characteristics, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, and shallow and deep water kelp populations have not been considered using advanced numerical and statistical tools.

The kelps *Laminaria pallida* Greville and *Ecklonia maxima* (Osbeck) Papenfuss are important habitat forming seaweeds that exist around the coast of South Africa and offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between canopy and sub-canopy species, as well as between deep and shallow water environments. Although both these species exist in the sub-tidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in the shallow sub-littoral. Therefore, *E. maxima* may be exposed to variations in wind driven surface waves, swell waves and temperature, while *L. pallida* is exposed to variations in swell waves and temperature. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. The authors suggested that turbidity in relation to light attenuation was the environmental driver responsible for this change.

The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology of sub-canopy and canopy kelps, as well as between deep and shallow populations of the same species. The current study will also aim to identify specific temperature and wave metrics which drive kelp morphological characteristics.  This will be achieved by initially understanding the variation in temperature and waves using an advanced numerical model. The numerical model will then be used as a basis for investigating the consequences for morphological characteristics of *E. maxima* and *L. pallida* using advanced statistical tools.

>>>>

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], form the basis of one of the most productive ecosystems globally [@Mann1973; @Krumhansl2012]. Kelps generally have a dependence on cool-temperate and arctic seawater temperatures [@Bolton2010; @Mohring2014-bk; @Cavanaugh2011-ib; @Dayton1985], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and sub-canopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Steneck2002], each depending on a wide suite of ecological services provided by kelp forests [@Dayton1985; @Gaines1987-bg; @Paul1993-bm; @Levin1994-qf; @Anderson1997; @Willis2003-xo]. Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in their distribution [@Gorman2013-nj], abundance [@Dayton1998; @Cavanaugh2011-ib], diversity [@Wing2007-tv; @Wernberg2008], composition [@Dayton1985; @Leliaert2000a; @Harley2012; @Norderhaug2012-ks], growth [@Cousens1982-ck] and productivity [@Dayton1998;@Krumhansl2012; @Pedersen2012].

Given that kelps are cool-temperate organisms which are vulnerable to dislodgement under high wave exposure scenarios, increases in storm frequency and magnitude, and changes in ocean temperature, pose a direct threat to their survival; which is a concern considering that these variables are known to be affected by climate change [@Russo2014; @Reguero2019; @Weisse2010; @Rose2012; @Rummukainen2012]. Warmer temperatures reduces the resilience of kelp individuals to storm disturbance [@Wernberg2010b], causes fragmentation through weakening tissues [@Simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist across a broad range of environments which is largely due to their morphological plasticity [@FowlerWalker2006; @Wernberg2010a]. A study by [@Wernberg2003-mp] investigated the morphology of Ecklonia radiata (C.Agardh) J.Agardh in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australian coast. They found no correlation between spatial distance and morphological similarity; rather the morphology of the kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp].

Although multiple environmental factors play a role in kelp morphological adaptation, wave exposure and temperature have been identified as the main influencers across various kelp species [@Hurd2000;@Denny2002;@Thomsen2004;@FowlerWalker2005;@Wernberg2005c;@mabin2013;@Bekkby2014]. For example, a study by  @Serisawa2002-d compared the morphology of Ecklonia cava Kjellman growing at warmer and cooler sites, and showed wrinkling of the blade and reduced size to be a characteristic of higher temperature ranges. Warmer temperature affects important kelp physiological processes such as photosynthesis and respiration, which influences growth and productivity and ultimately leads to a reduction in size [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of E. radiata across a broad geographic range, and showed trends toward drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of exposure. Cartographical models of wave exposure are “fetch-based-models” which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modeling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitative, reproducible approach for characterizing the hydrodynamic environment. Although @mabin2013 investigated kelp morphological characteristics using a 2D WAM model, currently no studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modeling. Furthermore no studies have identified specific wave metrics that drive kelp morphological characteristics, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, and shallow and deep water kelp populations have not been considered using advanced numerical and statistical tools. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001], and effects of wave dampening [@Kobayashi1993-os; @Mia2013; @Dubi1995].

The kelps *Laminaria pallida* Greville and *Ecklonia maxima* (Osbeck) Papenfuss are important habitat forming seaweeds that exist around the coast of South Africa and offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between canopy and sub-canopy species, as well as between deep and shallow water environments. Although both these species exist in the sub-tidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in the shallow sub-littoral. Therefore, E. maxima may be exposed to variations in wind driven surface waves, swell waves and temperature, while *L. pallida* is exposed to variations in swell waves, wave exposure and temperature due to differences in depth. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. The authors suggested that turbidity in relation to light attenuation was the environmental driver responsible for this change.

The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology of sub-canopy and canopy kelps, as well as between deep and shallow populations of the same species. The current study will also aim to identify specific temperature and wave metrics which drive kelp morphological characteristics. This will be achieved by initially understanding the variation in temperature and waves using a numerical model. The numerical model will then be used as a basis for investigating the consequences for morphological characteristics of *E. maxima* and *L. pallida* using advanced statistical tools.
